\name: A hierarchical Bayesian model for Supernova Cosmology

### 2018-02-26 papers/methods/paper.tex
Paper proof

% Abstract of the paper
\begin{abstract}
We present a hierarchical Bayesian model for use in supernova cosmology which takes into account full analysis systematics, Malmquist bias, underlying populations and standardisation coefficients. Our model is optimised for efficient fitting and has been validated on a combined dataset of more than $250\,000$ supernovae. Simplified statistical simulations for ideal supernovae cosmology show no bias, and sophisticated SNANA simulations quantify the bias for Flat $w$CDM with a matter prior $\Omega_m \sim \mathcal{N}(0.3, 0.01)$ allow rigorous determination of model bias in $w$ when provided with information on matter density (such as might be found from CMB constraints). Bias is found to range from $-0.15\sigma$ to $+0.27\sigma$ across intrinsic dispersion models for datasets of comparable strength to the Dark Energy Survey three-year analysis. 
\end{abstract}

@@

Almost two decades have passed since the discovery of the accelerating universe \citep{Riess1998, Perlmutter1999}. Since that time, of the number of observed Type Ia supernovae (SN Ia) have increased by more than an order of magnitude thanks to modern surveys at both low redshift \citep{Bailey2008, Freedman2009, Hicken2009,  Contreras2010, Conley2011}, and higher redshift \citep{Astier2006, Wood-Vasey2007, Balland2009, Amanullah2010}. Cosmological analysis of these supernova samples \citep{Kowalski2008, Conley2011, Suzuki2012, Betoule2014, Rest2014} have been combined with complimentary probes of large scale structure \citep{Alam2017} and the CMB \citep{Hinshaw2013, PlanckCollaboration2013}, and yet, despite these prodigious efforts, the nature of dark energy remains an unsolved mystery.

In attempts to tease out the nature of dark energy, currently running and planned surveys are once again ramping up their statistical power. The Dark Energy Survey \citep[DES,][]{Bernstein2012, Abbott2016} will be observing thousands of Type Ia supernova, attaining both spectroscopic and photometric confirmation. The Large Synoptic Survey Telescope \citep[LSST,][]{Ivezic2008, LSSTScienceCollaboration2009} will produce scores of thousands of photometrically classified supernovae. Such increased statistical power demands a similarly increased fidelity and flexibility in modelling the supernovae for cosmological purposes, as systematic uncertainty will prove to be the limiting factor in our analyses.

As such, staggering effort is being put into developing more sophisticated supernovae analyses. First used in ESSENSE analyses \citep{Wood-Vasey2007}, and then refined and improved in \citet{Kessler2009}, the role of simulations has become increasingly important in determining biases and validating models. \citet{Betoule2014} make use of redshift dependent $\mu$-bias to correct their analyses, and more recently \citet{Scolnic2016} and \citet{Kessler2017} explore simulation corrections as a function of redshift, stretch and colour to better remove fitting biases. Approximate Bayesian computation methods also make use of simulations, trading traditional likelihoods and analytic approximations for more robust models with only the cost of increased computational time \citep{Weyant2013, Jennings2016}. Hierarchical Bayesian Models abound \citep{Mandel2009, March2011, March2014, Rubin2015, Shariff2016, Roberts2017}, however often face difficulties finding sufficient analytic approximations for complicated effects such as Malmquist bias.

@@

Whilst supernova observations take the form of time-series photometric measurements of brightness in many photometric bands, most analyses do not work from these measurements of apparent magnitude and colour. Instead, most techniques fit these observations of magnitude (along with redshift) to a supernova model, with the most widely used being that of the empirical SALT2 model \citep{Guy2007, Guy2010}. This model is trained separately before fitting the supernovae light curves for the cosmology selected supernova sample \citep{Guy2010, Mosher2014}. The resulting output from the model is, for each supernova, a characterised amplitude $x_0$ (which can be converted into apparent magnitude $m_B = -2.5\log(x_0)$), a stretch term $x_1$ and colour term $c$, along with a covariance matrix describing the uncertainty on these summary statistics. As such, the product at the end is a (redshift-dependent) population of $m_B$, $x_1$ and $c$.

The underlying actual supernova population is not as clear cut, and indeed accurately characterising this population, its evolution over redshift and effects from environment is one of the challenges of supernova cosmology. However, given some modelled underlying population that lives in the redshift-dependent space $M_B$, $x_1$ and $c$, the introduction of cosmology into the model is simple -- it is encoded in the functional map between those two populations, from apparent magnitude space to absolute magnitude. Specifically, for any given supernova our functional map may take the traditional form:
\begin{equation}
M_B = m_B + \alpha x_1 - \beta c - \mu(z) + \text{corrections},
\end{equation}
where $\alpha$ is the stretch correction \citep{Phillips1993}, and $\beta$ is the colour correction \citep{Tripp1998} that respectively encapsulate the empirical relation that broader and bluer supernovae are brighter. The $\text{corrections}$ term at the end often includes corrections for host galaxy environment, as this has statistically significant effects on supernova properties \citep{Kelly2010, Lampeitl2010, Sullivan2010, DAndrea2011, Gupta2011, Johansson2013, Rigault2013, Uddin2017}. The cosmological term, $\mu(z)$ represents the distance modulus, and is precisely known given cosmological parameters and an input redshift.

@@

Traditional $\chi^2$ analyses such as that found in \citet{Kowalski2008, Conley2011, Betoule2014}, minimise the difference in distance modulus between the cosmologically predicted values $\mu_C$ and the observed distance modulus $\mu_{\rm obs}$, shown respectively below:
\begin{align}
\mu_C &= 5 \log\left[ \frac{(1+z)r}{10} \right]\\
r& =\frac{c}{H_0} \int_0^z \frac{dz'}{\sqrt{ \Omega_m (1+z')^3 + \Omega_k (1 + z')^2 + \Omega_\Lambda (1+z')^{3(1+w)}}} \label{eq:mu0}\\
\mu_{\rm obs} &= m_B + \alpha x_1 - \beta c - M_B \label{eq:mu1}
\end{align}
The minimising function is then given as
\begin{equation}
\chi^2 = (\mu_{\rm obs} - \mu_C)^\dagger C^{-1} (\mu_{\rm obs} - \mu_C)
\end{equation}
where $C^{-1}$ is an uncertainty matrix that combines the uncertainty from the SALT2 fits, intrinsic dispersion, calibration, dust, peculiar velocity and many other factors (see \citet{Betoule2014} for a review). The benefit this analysis methodology provides is speed - for samples of hundreds of supernovae, efficient matrix inversion algorithms allow the likelihood to be evaluated quickly. The speed comes with two costs. Firstly, formulating a $\chi^2$ likelihood requires a loss of model flexibility by building into the model assumptions of uncertainty Gaussianity. Secondly, the computational efficiency is dependent on inverting a covariance matrix with dimensionality linearly proportional to the number of supernovae. As this number increases, the cost of inversion rises quickly, and is not viable for samples with thousands of supernovae.

@@

Sitting comfortably between the traditional models simplicity and the complexity of forward modelling lies Hierarchical Bayesian Models. With the introduction of multiple layers in our model, we can add far more flexibility than a traditional analysis whilst still maintaining most of the computational benefits that come from having a tractable likelihood. \citet{Mandel2009}, \citet{Mandel2011} and \citet{Mandel2017} construct a hierarchical model that they apply to the light curve fitting for supernova. \citet{March2011} derive a hierarchical model and simplify it by analytically marginalising over nuisance parameters to provide a model that offers increased flexibility with reduced uncertainty over the traditional method, but does not incorporate bias correction. \citet{March2014, Karpenka2015} improve upon this by incorporating redshift-dependent magnitude corrections from \citet{Perrett2010} to remove bias, and validate on 100 realisations of SNLS like simulations. The recent BAHAMAS model \citep{Shariff2016} builds off this and reanalyses the JLA dataset (using their redshift dependent bias corrections), whilst including extra freedom in the correction factors $\alpha$ and $\beta$, finding evidence for redshift dependence on $\beta$. \citet{Ma2016} also performed a reanalysis of the JLA dataset, finding significant differences in $\alpha$ and $\beta$ values from the original as well by correctly including in their likelihood the determinant of the global covariance matrix. Notably, these methods rely on data that is bias corrected or ignore biases, however the UNITY framework given by \citet{Rubin2015} incorporates selection efficiency analytically in the model, and is applied to the Union 2.1 dataset \citep{Suzuki2012}. The simplification made by the UNITY analysis is that the bias is well described by an analytic function, and this function is fixed such that selection effects are assumed to be determined with perfect precision. Their model was validated on thirty realisations of statistics. Whilst there are a large amount of hierarchical models available, all sofar mentioned, in the case where any validation has been done, have been done on either $\Lambda$CDM cosmology or Flat $\Lambda$CDM cosmology. None of them have undergone high-statistics simulation verification for $w$CDM cosmology to quantify each models respective bias, which is becoming critically important as precision supernovae cosmology comes into its own and focus shifts from determination of $\Omega_m$ to $w$.

@@

We construct our Bayesian Hierarchical Model with several goals in mind: creation of a redshift-dependent correlated underlying supernova population, increased treatment of systematics, and analytic correction of selection effects. As this is closest to the UNITY method from \citet[][hereafter denoted \rubin]{Rubin2015}, we follow a similar model scaffold, and construct the model in Stan \citep{Carpenter2017, StanDevelopmentTeam2017} which uses automatic differentiation and the no-U-turn Sampler (NUTS), which is a variant of Hamiltonian Monte Carlo, to efficiently sample high dimensional parameter space.

@@

Like most of the BHM methods introduced previously, we work from the summary statistics as well, where each observed supernova has an apparent magnitude $\hat{m_B}$, stretch $\hat{x_1}$ and colour $\hat{c}$, with uncertainty $\cov$ on those values. Additionally, each supernova has an observed redshift $\hat{z}$ and a host galaxy mass associated with it, $\hat{m}$, where the mass measurement takes the form of a probability of being above $10^10$ solar masses. Given machine learning classifiers to type supernovae, we will also have a probability of being a Type Ia, $\hat{p}$. Our set of observables is therefore given as $\lbrace \hat{m_B}, \hat{x_1}, \hat{c}, \hat{z}, \hat{p}, \hat{m}, \cov \rbrace$, as shown in the probabilistic graphical model (PGM) in Figure \ref{fig:pgm}.

@@

The underlying supernova population is often treated with two components - a population distribution in colour and stretch, and intrinsic dispersion. Analytic models often treat the colour and stretch populations with a skew normal and normal, respectively, and have them as independent. Intrinsic dispersion is treated in a variety of manners in other models, from representing it simply as (Gaussian) scatter on the absolute magnitude of the supernova population, to correlated multivariate normal scatter on the combined magnitude, stretch and colour distribution. Additionally, redshift drift of populations' mean colour and stretch will introduce a cosmological bias in our fits unless the population possesses similar ability to change as a function of redshift.

@@

As such, the only constant between survey populations is the absolute magnitude $M_B$, with the skewness, redshift-dependent means and width fit individually for each survey. The probability for a supernova to have true values $M_B$, $x_1$, $c$ given the underlying population is thus given as
\begin{align}
P(M_B, x_1, c, z | \langle M_B \rangle, \langle x_1(z) \rangle, \langle c(z) \rangle, \rho, \sigma_{m_B}, \sigma_{x_1}, \sigma_c, \alpha_c) = \notag \\
\mathcal{N}(M_B|\langle M_B \rangle, \sigma_{M_B}) \mathcal{N}(x_1 | \langle x_1(z) \rangle \sigma_{x_1}) \mathcal{N}^{\rm{skew}}(c| \langle c(z) \rangle, \sigma_c, \alpha_c) \label{eq:l1}.
\end{align}

@@

With increasingly large datasets and more nuanced analyses, the choice of how to handle $\alpha$ and $\beta$ becomes an important consideration when constructing a model. {\rubin} employs a broken linear relationship for both colour and stretch, where different values of $\alpha$ and $\beta$ are adopted depending on whether $x_1$ and $c$ are respectively positive or negative (although the cut could be placed at a location other than 0). \citet{Shariff2016} instead of employ a colour-dependent $\beta$, model $\beta$ as redshift-dependent, testing two phenomenological models; $\beta(z) = \beta_0 + \beta_1 z$ and $\beta(z) = \beta_0 + \Delta \beta\left(0.5 + \arctan(100(z - z_t)) / \pi\right)$, where the later effects a rapid but smooth change in $\beta$ at a turnover redshift $z_t$.

@@

Our treatment of selection effects is to incorporate selection efficiency into our model, rather than relying on simulation-driven data corrections. As such, we need to describe the probability that the events we observe are both drawn from the distribution predicted by the underlying theoretical model \textit{and} that those events, given they happened, are subsequently successfully observed.  To make this extra conditional explicit, we can write the likelihood of the data given an underlying model, $\theta$, \textit{and} that the data are included in our sample, denoted by $S$, as

@@

where the denominator represents an integral over all potential data. Full derivation of this can be found in Appendix \ref{app:selection}. As $\theta$ represents the vector of all model parameters, and $D$ represents a vector of all observed variables, this is not a trivial integral. Techniques to approximate this integral, such as Monte-Carlo integration or high dimensional Gaussian processes failed to give tractable posterior surfaces that could be sampled efficiently by HMC (a brief dismissal of months of struggle). We therefore simplify the integral and approximate the selection effects in apparent magnitude and redshift space independently, such that the denominator, denoted now $w$ for simplicity, is given as

@@

We apply two further approximations similar to those made in {\rubin} -- that the redshift distribution of the observed supernova reasonably well sampled the $P(S|z)P(z|\theta)$ distribution, and that the survey colour and stretch populations can be treated as Gaussian for the purposes of this integral. It was found that discarding the skewness entirely resulted in highly biased population recovery, and so we instead characterise the skew normal colour distribution with a Gaussian that follows the mean and variance of a skew normal, with mean given by $\langle c(z) \rangle + \sqrt{\frac{2}{\pi}} \sigma_c \delta_c$ and variance $\sigma_c^2(1 - 2\delta_c^2/\pi)$. This shifted Gaussian approximation removes the unintended bias when simply discarding skewness. More detail on this shift can be found in Appendix \ref{app:approx}. The population $P(m_B | z, \theta)$ becomes $\mathcal{N}(m_B|m_B^*(z), \sigma^*_{m_B})$, where

@@

where $\Phi^c$ the complimentary CDF and $\mu_{\rm CDF}$ and $\sigma_{\rm CDF}$ specify the selection function. The appropriateness of an error function has been found by many past surveys \citep{Dilday2008, Barbary2010, Perrett2012, Graur2013, Rodney2014}. However, for surveys which suffer from saturation and thus rejection of low-z supernovae, or for groups of surveys treated together (as is common to do with low-redshift surveys), we find that a skew normal is a good analytic form, taking the form

@@

The selection functions are fit to apparent magnitude efficiency ratios calculated from SNANA simulations, by calculating an efficiency ratio as a function of apparent magnitude. Uncertainty of the Malmquist bias (entering both through statistical uncertainty from finite sized simulations in the efficiency ratio and the discrepancy between the analytic approximation and non-analytic simulation results) is incorporated into the fitting for the analytic approximation. Uncertainty is uniformly added to the efficiency ratio until the reduced $\chi^2$ of the analytic fit reached $1$, allowing us to extract an uncertainty covariance matrix for our analytic fits to either the error function or the skew normal.

With the well sampled approximation as specified previously, we can remove the redshift integral in Eq \eqref{eq:w1} and replace it with a correction for each observed supernova. For the error function (denoted with the subscript `CDF') and skew normal selection functions respectively (denoted with a subscript `Skew'), this correction becomes
\begin{align}
w_{\rm CDF} &= \Phi^c\left(  \frac{m^*_B - \mu_{\rm CDF}}{\sqrt{\sigma^{*2}_{m_B} + \sigma_{\rm CDF}^2}}  \right) \label{eq:seldes}\\
w_{\rm Skew} &= 2\mathcal{N}\left( \frac{m_B^* - \mu_{\rm Skew}}{\sqrt{\sigma^{*2}_{m_B} + \sigma_{\rm Skew}^2}}\right) \times \notag\\
&\quad\quad \Phi\left(  \frac{{\rm sign}(\alpha_{\rm Skew})(m_B^* - \mu_{\rm Skew})}{\frac{\sigma_{m_B}^{*2} + \sigma^2_{\rm Skew}}{\sigma^2_{\rm Skew}} \sqrt{\frac{\sigma^2_{\rm Skew}}{\alpha^2_{\rm Skew}} + \frac{\sigma_{m_B}^{*2} \sigma^2_{\rm Skew}}{\sigma_{m_B}^{*2} + \sigma^2_{\rm Skew}}} }\right). \label{eq:sellowz}
\end{align}

@@

We generate simple toy data to validate the basic premise of the model. For both high-redshift and low-redshift (LowZ) data we draw from an underlying $M_B$, $x_1$, $c$ population and translate into apparent magnitude space using $\Omega_m = 0.3$, $\alpha= 0.14$ and $\beta = 3.1$. Masses are randomly drawn from the interval 0 to 1, and a mass correction with $\delta(0) = 0.08$ and $\delta(0)/\delta(\infty) =0.5$ included. Absolute magnitudes are drawn from $\mathcal{N}(-19.3, 0.1)$, true stretch values are drawn from $\mathcal{N}(0, 1)$ and true colour values are drawn from $\mathcal{N}^{\rm Skew}(0, 0.1, 2)$. Redshifts are drawn from a power law to increase the number of events as redshift increases.

Independent observational errors of $0.04$, $0.2$, $0.03$ on $m_B$, $x_1$ and $c$ (following the mean uncertainty for DES SNANA simulations) are added to create the observables. Extra dispersion of $0.02(1+3z)$ is added in quadrature to the $c$ uncertainty to simulation unexplained colour dispersion from chromatic variation in supernovae. The selection functions (a skew normal for low-redshift and an error function for high-redshift) are given independent uncertainty of $0.01$ on all parameters. The results are then passed through the selection effects, where each supernova is only selected based on $P(S|m_B)$, using a skew normal function for the LowZ supernovae and error function for the DES-like supernovae. We draw from each survey simulation until we have 1000 LowZ supernovae and 1000 DES-like supernovae, representing a statistical sample of greater power than the estimated 250 supernovae for the DES spectroscopic analysis. Sample data for 1000 high and low redshift supernovae are shown in Figure \ref{fig:simple_pop}, confirming the presence of strong selection effects in both toy surveys.

>>>>

% Abstract of the paper
\begin{abstract}
We present a hierarchical Bayesian model for use in supernova cosmology which takes into account full analysis systematics, Malmquist bias, underlying populations and standardisation coefficients. Our model is optimised for efficient fitting and has been validated on a combined dataset of more than $250\,000$ supernovae. Simplified statistical simulations for ideal supernovae cosmology show no bias, and sophisticated SNANA simulations quantify the bias for Flat $w$CDM cosmology, with a matter prior $\Omega_m \sim \mathcal{N}(0.3, 0.01)$ included to allow rigorous determination of model bias in $w$. Bias in SNANA tests across intrinsic dispersion models is found to range from $-0.15\sigma$ to $+0.27\sigma$ in $w$ for datasets of comparable strength to the Dark Energy Survey three-year analysis. 
\end{abstract}

@@

Almost two decades have passed since the discovery of the accelerating universe \citep{Riess1998, Perlmutter1999}. Since that time, of the number of observed Type Ia supernovae (SN Ia) have increased by more than an order of magnitude thanks to modern surveys at both low redshift \citep{Bailey2008, Freedman2009, Hicken2009,  Contreras2010, Conley2011}, and higher redshift \citep{Astier2006, Wood-Vasey2007, Balland2009, Amanullah2010}. Cosmological analysis of these supernova samples \citep{Kowalski2008, Conley2011, Suzuki2012, Betoule2014, Rest2014} have been combined with complimentary probes of large scale structure \citep{Alam2017} and the CMB \citep{Hinshaw2013, PlanckCollaboration2013}. Despite these prodigious efforts, the nature of dark energy remains an unsolved mystery.

In attempts to tease out the nature of dark energy, active and planned surveys are once again ramping up their statistical power. The Dark Energy Survey \citep[DES,][]{Bernstein2012, Abbott2016} will be observing thousands of Type Ia supernova, attaining both spectroscopic and photometric confirmation. The Large Synoptic Survey Telescope \citep[LSST,][]{Ivezic2008, LSSTScienceCollaboration2009} will produce scores of thousands of photometrically classified supernovae. Such increased statistical power demands increased fidelity and flexibility in modelling the supernovae for cosmological purposes, as systematic uncertainty will prove to be the limiting factor in our analyses.

As such, staggering effort is being put into developing more sophisticated supernovae analyses. First used in ESSENSE analyses \citep{Wood-Vasey2007}, and then refined and improved in \citet{Kessler2009}, the role of simulations has become increasingly important in determining biases and validating models. \citet{Betoule2014} make use of redshift dependent $\mu$-bias to correct their analyses, and more recently \citet{Scolnic2016} and \citet{Kessler2017} explore simulation corrections as a function of redshift, stretch and colour to better remove fitting biases. Approximate Bayesian computation methods also make use of simulations, trading traditional likelihoods and analytic approximations for more robust models with the cost of increased computational time \citep{Weyant2013, Jennings2016}. Hierarchical Bayesian Models abound \citep{Mandel2009, March2011, March2014, Rubin2015, Shariff2016, Roberts2017}, however often face difficulties finding sufficient analytic approximations for complicated effects such as Malmquist bias.

@@

Whilst supernova observations take the form of time-series photometric measurements of brightness in many photometric bands, most analyses do not work from these measurements. Instead, most techniques fit these observations (along with redshift) to a supernova model, with the most widely used being that of the empirical SALT2 model \citep{Guy2007, Guy2010}. This model is trained separately before fitting the supernovae light curves for the cosmology selected supernova sample \citep{Guy2010, Mosher2014}. The resulting output from the model is, for each supernova, a characterised amplitude $x_0$ (which can be converted into apparent magnitude $m_B = -2.5\log(x_0)$), a stretch term $x_1$ and colour term $c$, along with a covariance matrix describing the uncertainty on these summary statistics. As such, the product at the end is a (redshift-dependent) population of $m_B$, $x_1$ and $c$.

The underlying actual supernova population is not as clear cut, and indeed accurately characterising this population, its evolution over redshift and effects from environment is one of the challenges of supernova cosmology. However, given some modelled underlying population that lives in the redshift-dependent space $M_B$, $x_1$ and $c$, the introduction of cosmology into the model is simple -- it is encoded in the functional map between those two populations, from apparent magnitude space to absolute magnitude. Specifically, for any given supernova our functional map may take the traditional form:
\begin{equation}
M_B = m_B + \alpha x_1 - \beta c - \mu(z) + \text{corrections}, \label{eq:standard}
\end{equation}
where $\alpha$ is the stretch correction \citep{Phillips1993}, and $\beta$ is the colour correction \citep{Tripp1998} that respectively encapsulate the empirical relation that broader and bluer supernovae are brighter. The $\text{corrections}$ term at the end often includes corrections for host galaxy environment, as this has statistically significant effects on supernova properties \citep{Kelly2010, Lampeitl2010, Sullivan2010, DAndrea2011, Gupta2011, Johansson2013, Rigault2013, Uddin2017}. The cosmological term, $\mu(z)$ represents the distance modulus, and is precisely known given cosmological parameters and an input redshift.

@@

Traditional $\chi^2$ analyses such as that found in \citet{Kowalski2008, Conley2011, Betoule2014}, minimise the difference in distance modulus between the observed distance modulus $\mu_{\rm obs}$ (attained from rearranging equation \eqref{eq:standard}), and the cosmologically predicted values $\mu_C$; given by
\begin{align}
\mu_C &= 5 \log\left[ \frac{(1+z)r}{10} \right]\\
r& =\frac{c}{H_0} \int_0^z \frac{dz'}{\sqrt{ \Omega_m (1+z')^3 + \Omega_k (1 + z')^2 + \Omega_\Lambda (1+z')^{3(1+w)}}} \label{eq:mu0}\\
\mu_{\rm obs} &= m_B + \alpha x_1 - \beta c - M_B. \label{eq:mu1}
\end{align}
The minimising function is then given as
\begin{equation}
\chi^2 = (\mu_{\rm obs} - \mu_C)^\dagger C^{-1} (\mu_{\rm obs} - \mu_C)
\end{equation}
where $C^{-1}$ is an uncertainty matrix that combines the uncertainty from the SALT2 fits, intrinsic dispersion, calibration, dust, peculiar velocity and many other factors (see \citet{Betoule2014} for a review). The benefit this analysis methodology provides is speed - for samples of hundreds of supernovae, efficient matrix inversion algorithms allow the likelihood to be evaluated quickly. The speed comes with two costs. Firstly, formulating a $\chi^2$ likelihood requires a loss of model flexibility by building into the model assumptions of uncertainty Gaussianity. Secondly, the method of creating a global covariance matrix relies on computing partial derivatives and thus loses all information about uncertainty correlation. Thirdly, the computational efficiency is dependent on inverting a covariance matrix with dimensionality linearly proportional to the number of supernovae. As this number increases, the cost of inversion rises quickly, and is not viable for samples with thousands of supernovae.

@@

Sitting comfortably between the traditional models simplicity and the complexity of forward modelling lies hierarchical Bayesian Models. With the introduction of multiple layers in our model, we can add far more flexibility than a traditional analysis whilst still maintaining most of the computational benefits that come from having a tractable likelihood. \citet{Mandel2009}, \citet{Mandel2011} and \citet{Mandel2017} construct a hierarchical model that they apply to the light curve fitting for supernova. \citet{March2011} derive a hierarchical model and simplify it by analytically marginalising over nuisance parameters to provide a model that offers increased flexibility with reduced uncertainty over the traditional method, but does not incorporate bias correction. \citet{March2014, Karpenka2015} improve upon this by incorporating redshift-dependent magnitude corrections from \citet{Perrett2010} to remove bias, and validate on 100 realisations of SNLS like simulations. The recent BAHAMAS model \citep{Shariff2016} builds off this and reanalyses the JLA dataset (using their redshift dependent bias corrections), whilst including extra freedom in the correction factors $\alpha$ and $\beta$, finding evidence for redshift dependence on $\beta$. \citet{Ma2016} also performed a reanalysis of the JLA dataset, finding significant differences in $\alpha$ and $\beta$ values from the original as well by correctly including in their likelihood the determinant of the global covariance matrix. Notably, these methods rely on data that is bias corrected or ignore biases, however the UNITY framework given by \citet{Rubin2015} incorporates selection efficiency analytically in the model, and is applied to the Union 2.1 dataset \citep{Suzuki2012}. The simplification made by the UNITY analysis is that the bias is well described by an analytic function, and this function is fixed such that selection effects are assumed to be determined with perfect precision. Their model was validated on thirty realisations of statistics. Whilst there are a large amount of hierarchical models available, all sofar mentioned, in the case where any validation has been done, have been done on either $\Lambda$CDM cosmology or Flat $\Lambda$CDM cosmology. None of them have undergone high-statistics simulation verification for $w$CDM cosmology to quantify each models respective bias, which is becoming critically important as precision supernovae cosmology comes into its own and focus shifts from determination of $\Omega_m$ to $w$.

@@

We construct our hierarchical Bayesian model with several goals in mind: creation of a redshift-dependent underlying supernova population, increased treatment of systematics, and analytic correction of selection effects, including systematic uncertainty. As this is closest to the UNITY method from \citet[][hereafter denoted \rubin]{Rubin2015}, we follow a similar model scaffold, and construct the model in Stan \citep{Carpenter2017, StanDevelopmentTeam2017} which uses automatic differentiation and the no-U-turn Sampler (NUTS), which is a variant of Hamiltonian Monte Carlo, to efficiently sample high dimensional parameter space.

@@

Like most of the BHM methods introduced previously, we work from the summary statistics as well, where each observed supernova has an apparent magnitude $\hat{m_B}$, stretch $\hat{x_1}$ and colour $\hat{c}$, with uncertainty $\cov$ on those values. Additionally, each supernova has an observed redshift $\hat{z}$ and a host galaxy mass associated with it, $\hat{m}$, where the mass measurement takes the form of a probability of being above $10^{10}$ solar masses. Given machine learning classifiers to type supernovae, we will also have a probability of being a Type Ia, $\hat{p}$. Our set of observables is therefore given as $\lbrace \hat{m_B}, \hat{x_1}, \hat{c}, \hat{z}, \hat{p}, \hat{m}, \cov \rbrace$, as shown in the probabilistic graphical model (PGM) in Figure \ref{fig:pgm}.

@@

The underlying supernova population is characterised by underlying distributions in colour, stretch and magnitude. Analytic models often treat the colour and stretch populations with a skew normal and normal, respectively, and have them as independent. Intrinsic dispersion is treated in a variety of manners in other models, from representing it simply as (Gaussian) scatter on the absolute magnitude of the supernova population, to correlated multivariate normal scatter on the combined magnitude, stretch and colour distribution. Additionally, redshift drift of populations' mean colour and stretch will introduce a cosmological bias in our fits unless the population possesses similar ability to change as a function of redshift.

@@

As such, the only constant between survey populations is the absolute magnitude $M_B$, with the skewness, redshift-dependent means and width fit individually for each survey. The probability for a supernova to have true values $M_B$, $x_1$, $c$ given the underlying population is thus given as
\begin{align}
P(M_B, x_1, c, z | \langle M_B \rangle, \langle x_1(z) \rangle, \langle c(z) \rangle, \rho, \sigma_{m_B}, \sigma_{x_1}, \sigma_c, \alpha_c) = \notag \\
\mathcal{N}(M_B|\langle M_B \rangle, \sigma_{M_B}) \mathcal{N}(x_1 | \langle x_1(z) \rangle, \sigma_{x_1}) \mathcal{N}^{\rm{skew}}(c| \langle c(z) \rangle, \sigma_c, \alpha_c) \label{eq:l1}.
\end{align}

@@

With increasingly large datasets and more nuanced analyses, the choice of how to handle $\alpha$ and $\beta$ becomes an important consideration when constructing a model. {\rubin} employs a broken linear relationship for both colour and stretch, where different values of $\alpha$ and $\beta$ are adopted depending on whether $x_1$ and $c$ are respectively positive or negative (although the cut could be placed at a location other than 0). \citet{Shariff2016} instead model $\beta$ as redshift-dependent, testing two phenomenological models; $\beta(z) = \beta_0 + \beta_1 z$ and $\beta(z) = \beta_0 + \Delta \beta\left(0.5 + \arctan(100(z - z_t)) / \pi\right)$, where the later effects a rapid but smooth change in $\beta$ at a turnover redshift $z_t$.

@@

Our treatment of selection effects is to incorporate selection efficiency into our model, rather than relying on simulation-driven data corrections. We need to describe the probability that the events we observe are both drawn from the distribution predicted by the underlying theoretical model \textit{and} that those events, given they happened, are subsequently successfully observed.  To make this extra conditional explicit, we can write the likelihood of the data given an underlying model, $\theta$, \textit{and} that the data are included in our sample, denoted by $S$, as

@@

where the denominator represents an integral over all potential data. Full derivation of this can be found in Appendix \ref{app:selection}. As $\theta$ represents the vector of all model parameters, and $D$ represents a vector of all observed variables, this is not a trivial integral. Techniques to approximate this integral, such as Monte-Carlo integration or high dimensional Gaussian processes failed to give tractable posterior surfaces that could be sampled efficiently by HMC, and post-fitting importance sampling failed due to high-dimensionality (a brief dismissal of many months of struggle). We therefore simplify the integral and approximate the selection effects in apparent magnitude and redshift space independently, such that the denominator of equation \eqref{eq:main}, denoted now $w$ for simplicity, is given as

@@

We apply two further approximations similar to those made in {\rubin} -- that the redshift distribution of the observed supernova reasonably well sampled the $P(S|z)P(z|\theta)$ distribution, and that the survey colour and stretch populations can be treated as Gaussian for the purposes of this integral. It was found that discarding the skewness entirely resulted in highly biased population recovery, and so we instead characterise the skew normal colour distribution with a Gaussian that follows the mean and variance of a skew normal; with mean given by $\langle c(z) \rangle + \sqrt{\frac{2}{\pi}} \sigma_c \delta_c$ and variance $\sigma_c^2(1 - 2\delta_c^2/\pi)$. This shifted Gaussian approximation removes the unintended bias when simply discarding skewness. More detail on this shift can be found in Appendix \ref{app:approx}. The population $P(m_B | z, \theta)$ becomes $\mathcal{N}(m_B|m_B^*(z), \sigma^*_{m_B})$, where

@@

where $\Phi^c$ the complimentary CDF and $\mu_{\rm CDF}$ and $\sigma_{\rm CDF}$ specify the selection function. The appropriateness of an error function has been found by many past surveys \citep{Dilday2008, Barbary2010, Perrett2012, Graur2013, Rodney2014}. However, for surveys which suffer from saturation and thus rejection of low-z supernovae, or for groups of surveys treated together (as is common to do with low-redshift surveys), we find that a skew normal is a suitable analytic form, taking the form

@@

The selection functions are fit to apparent magnitude efficiency ratios calculated from SNANA simulations, by calculating an efficiency ratio as a function of apparent magnitude. Uncertainty of the Malmquist bias from the imperfection of the analytic function is incorporated into the fit. Uncertainty was uniformly added to the efficiency ratio data from our simulations until the reduced $\chi^2$ of the analytic fit reached $1$, allowing us to extract an uncertainty covariance matrix for our analytic fits to either the error function or the skew normal.

With the well sampled approximation as specified previously, we can remove the redshift integral in Eq \eqref{eq:w1} and replace it with a correction for each observed supernova. For the error function (denoted with the subscript `CDF') and skew normal selection functions respectively (denoted with a subscript `Skew'), this correction becomes
\begin{align}
w_{\rm CDF} &= \Phi^c\left(  \frac{m^*_B - \mu_{\rm CDF}}{\sqrt{\sigma^{*2}_{m_B} + \sigma_{\rm CDF}^2}}  \right) \label{eq:seldes}\\
w_{\rm Skew} &= 2\mathcal{N}\left( \frac{m_B^* - \mu_{\rm Skew}}{\sqrt{\sigma^{*2}_{m_B} + \sigma_{\rm Skew}^2}}\right) \times \notag\\
&\quad\quad \Phi\left(  \frac{{\rm sign}(\alpha_{\rm Skew})(m_B^* - \mu_{\rm Skew})}{\frac{\sigma_{m_B}^{*2} + \sigma^2_{\rm Skew}}{\sigma^2_{\rm Skew}} \sqrt{\frac{\sigma^2_{\rm Skew}}{\alpha^2_{\rm Skew}} + \frac{\sigma_{m_B}^{*2} \sigma^2_{\rm Skew}}{\sigma_{m_B}^{*2} + \sigma^2_{\rm Skew}}} }\right), \label{eq:sellowz}
\end{align}
and is incorporated into our likelihood.

@@

We generate simple toy data to validate the basic premise of the model. For both high-redshift and low-redshift (LowZ) data we draw from an underlying $M_B$, $x_1$, $c$ population and translate into apparent magnitude space using $\Omega_m = 0.3$, $\alpha= 0.14$ and $\beta = 3.1$. Mass probabilities are randomly drawn from the interval 0 to 1, and a mass correction with $\delta(0) = 0.08$ and $\delta(0)/\delta(\infty) =0.5$ included. Absolute magnitudes are drawn from $\mathcal{N}(-19.3, 0.1)$, true stretch values are drawn from $\mathcal{N}(0, 1)$ and true colour values are drawn from $\mathcal{N}^{\rm Skew}(0, 0.1, 2)$. Redshifts are drawn from a power law to increase the number of events as redshift increases.

Independent observational errors of $0.04$, $0.2$, $0.03$ on $m_B$, $x_1$ and $c$ (following the mean uncertainty for DES SNANA simulations) are added to create the observables. Extra dispersion of $0.02(1+3z)$ is added in quadrature to the $c$ uncertainty to simulation unexplained colour dispersion from chromatic variation in supernovae. The selection functions (a skew normal for low-redshift and an error function for high-redshift) are given independent uncertainty of $0.01$ on all parameters. The results are then passed through the selection effects, where each supernova is only selected based on $P(S|m_B)$, using a skew normal function for the LowZ supernovae and error function for the DES-like supernovae. We draw from each survey simulation until we have 1000 LowZ supernovae and 1000 DES-like supernovae, representing a statistical sample of greater power than the estimated 350 supernovae for the DES spectroscopic analysis. Sample data for 1000 high and low redshift supernovae are shown in Figure \ref{fig:simple_pop}, confirming the presence of strong selection effects in both toy surveys.
